The Importance of Being Nonplanar: Street Network Representation in Urban Form Studies

### 2018-01-16 manuscript.tex
edits

In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently forbids overpasses or underpasses. Of course in the real-world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels. This leads to two intertwined questions underexplored in the literature today. First, how well do planar graphs model urban street networks? Second, what does the extent to which a network is nonplanar tell us about the city's infrastructure and development?

@@

This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}. A street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real-world), but it could still be formally planar. For example, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}. In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.

@@

Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street network research? How close is \enquote{close enough} for a planar graph to competently model a formally nonplanar street network? Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different degrees? And if street networks are at least generally approximately planar, then how can we measure just how planar or nonplanar a given street network is?

The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}. However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}. In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.

@@

Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide. This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents. For cities, such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center. We look separately at each city's drivable and walkable street networks.

@@

Once we have acquired and prepared our networks, we calculate three measures of planarity.

The first is an aspatial test of formal planarity using the algorithm described by \citet{boyer_subgraph_2012}. That is, is it possible to rearrange the graph's nodes and edges in space --- while preserving its topology --- so that edges cross only at nodes? This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding. However, street networks are spatially embedded, so we develop the following two continuous indicators to assess the \enquote{extent} to which they are planar.

The second measure is the spatial planarity ratio (SPR). It represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph). It indicates the extent to which planarity overstates intersections and connectivity in each street network. A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the extent to which the network is nonplanar.

The third measure is the edge length ratio (ELR). It represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph. It indicates the extent to which planarity understates edge lengths in each street network. A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the extent to which the network is nonplanar.

@@

\begin{figure}[htbp]
	\includegraphics[width=0.5\textwidth]{regression_phi_split.png}
	\caption{Log-log plot of ELR vs SPR with simple regression lines.}
	\label{fig:regression_split}
\end{figure}

@@

We find that the street network of the entire city of Oakland is formally nonplanar when subjected to a formal mathematical test. In terms of the nonplanarity indicators, the city as a whole has a SPR score of 0.918, indicating that it is 91.8\% planar, and a ELR score of 0.936. This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections (and thus, the network's connectivity) by 8.9\% city-wide (i.e., $1 - 1 / 0.918$ and understates the average edge length by 6.4\% city-wide (i.e., $1 - 0.936$). However, these indicators' values vary across the city.

@@

The urban studies and urban physics literatures frequently use planar graphs to urban street networks. This study demonstrated that although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths. It found x, y, and z. It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular its transportation infrastructure's three-dimensionality, of different cities. Future research can explore this latter finding, as it likely correlates with other measures of urbanization, development, and era. Finally, future research might examine how nonplanar intersection counts represent true intersections if multiple adjacent edges form multiple graph intersections at a point where only one true intersection exists from an urban design perspective.

>>>>

In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently forbids overpasses or underpasses. Of course in the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels. This leads to two intertwined questions. First, how well do these planar graphs model urban street networks? Second, what does the extent to which a network is nonplanar tell us about a city's infrastructure and development?

@@

This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}. For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar. If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}. In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.

@@

Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street network research? How close is \enquote{close enough} for a planar graph to competently model a formally nonplanar street network? Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different degrees? And if street networks are at least generally approximately planar, how can we measure just how planar or nonplanar a given street network is?

The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}. However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}. They also fail to account for the size or density of a spatial network. In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.

@@

Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide. This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents. For cities such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center. We look separately at each city's drivable and walkable street networks.

@@

Once we have acquired and prepared our networks, we calculate three measures of planarity. The first is an aspatial test of formal planarity using the algorithm described by \citet{boyer_subgraph_2012}. That is, is it possible to rearrange the graph's nodes and edges in space --- while preserving its topology --- so that edges cross only at nodes? This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding. However, street networks are spatially embedded, so we develop the following two continuous indicators to assess the \enquote{extent} to which they are planar.

The second measure is the spatial planarity ratio (SPR). It represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph). This indicates the extent to which planarity overstates intersections and connectivity in each street network. A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the extent to which the network is nonplanar.

The third measure is the edge length ratio (ELR). It represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph. This indicates the extent to which planarity understates edge lengths in each street network. A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the extent to which the network is nonplanar.

@@

\begin{figure}[htbp]
	\includegraphics[width=0.48\textwidth]{regression_phi_split.png}
	\caption{Log-log plot of ELR vs SPR with simple regression lines.}
	\label{fig:regression_split}
\end{figure}

@@

We find that the street network of the entire city of Oakland is formally nonplanar when subjected to a formal mathematical test. In terms of the nonplanarity indicators, the city as a whole has an SPR score of 0.918, indicating that it is 91.8\% planar, and an ELR score of 0.936. This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections (and thus, the network's connectivity) by 8.9\% city-wide (i.e., $1 - 1 / 0.918$ and understates the average edge length by 6.4\% city-wide (i.e., $1 - 0.936$). However, these indicators' values vary across the city.

@@

Urban researchers frequently use planar graphs to model street networks. This study demonstrated that, although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths. It found x, y, and z. It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular transportation infrastructure's three-dimensionality, in different cities. Future research can explore this latter finding, as it likely correlates with other measures of urbanization, development, and era. Finally, future research might examine how nonplanar intersection counts represent true intersections if multiple adjacent edges form multiple graph intersections at a point where only one true intersection exists from an urban design perspective.
